---
title: "Forecasting 10-year US treasury bond yields from PCE inflation rate"
author: "Christopher C. Smith"
date: '2022-06-18'
output: pdf_document
---

```{r setup, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
#Script to predict 10-year yields from PCE inflation rate, factoring in how 
#10-year yields have fallen over time

library(tidyverse)
library(xlsx)
library(lubridate)
library(RSelenium)

#Start Selenium driver and navigate to FRED series for 10-year yields
rD <- rsDriver(browser = "firefox")
driver <- rD[["client"]]
driver$navigate("https://fred.stlouisfed.org/series/DGS10")
Sys.sleep(2)


#Find "Max" button on the page and click it
element <- driver$findElement(using = "xpath",'//*[@id="zoom-all"]')
element$clickElement()
Sys.sleep(1)

#Find "Download" button on the page and click it
element <- driver$findElement(using = "css",'.fg-download-btn-chart-gtm > span:nth-child(1)')
element$clickElement()
Sys.sleep(1)

#Find "CSV" button on the page and get the link from it
element <- driver$findElement(using = "xpath",'//*[@id="download-data-csv"]')
url <- unlist(element$getElementAttribute('href'))

#Get and clean 10-year yield data
download.file(url,"DGS10.csv")
tenyr_df <- read.csv("DGS10.csv") %>%
  mutate(yield = as.numeric(DGS10)) %>%
  filter(!is.na(yield))
tenyr_df <- tenyr_df %>% 
  mutate(year = year(DATE),month = month(DATE)) %>%
  group_by(year,month) %>%
  summarize(mean_yield = mean(yield))
unlink("DGS10.csv")

#Navigate to FRED series for PCE inflation rate
driver$navigate("https://fred.stlouisfed.org/series/PCETRIM12M159SFRBDAL")
Sys.sleep(2)

#Find "Max" button on the page and click it
element <- driver$findElement(using = "xpath",'//*[@id="zoom-all"]')
element$clickElement()
Sys.sleep(1)

#Find "Download" button on the page and click it
element <- driver$findElement(using = "css",'.fg-download-btn-chart-gtm > span:nth-child(1)')
element$clickElement()
Sys.sleep(1)

#Find "CSV" button on the page and get the link from it
element <- driver$findElement(using = "xpath",'//*[@id="download-data-csv"]')
url <- unlist(element$getElementAttribute('href'))

#Get and clean PCE inflation dataset
download.file(url,"PCETRIM12M159SFRBDAL.csv")
pce_df <- read.csv("PCETRIM12M159SFRBDAL.csv") 
pce_df <- pce_df %>%
  mutate(year = year(DATE),month = month(DATE)) %>%
  mutate(twelve_month_pce = as.numeric(PCETRIM12M159SFRBDAL)) %>%
  filter(!is.na(twelve_month_pce)) %>%
  select(-DATE,-PCETRIM12M159SFRBDAL)
unlink("PCETRIM12M159SFRBDAL.csv")

#Join the two tables
joined_df <- inner_join(tenyr_df,pce_df) %>% 
  ungroup() %>%
  mutate(date = as.Date(paste(as.character(year),as.character(month),"1",sep="/"),format="%Y/%m/%d"),
         sequence = row_number())

#Predict 10-year yield based on this simple log regression model using PCE rate
fit <- joined_df %>% lm(mean_yield ~ log(twelve_month_pce), data = .)
initial_predicted_yield <- predict(fit,newdata=data.frame(twelve_month_pce = last(joined_df$twelve_month_pce)))
```

## Introduction

Versions of this chart have been making the rounds. I've shared it myself. Plot historical 10-year bond yields (Board of Governors, 2022a) against historical PCE inflation rates (Board of Governors, 2022b), run a simple log regression, and *voila!* today's 10-year yield should be `r paste(round(initial_predicted_yield,2),"%",sep="")`!

But this is not quite right.

```{r chart_1, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
#Plot PCE rate against 10-year yield, with date shown as color
joined_df %>%
  ggplot(aes(x=twelve_month_pce,y=mean_yield,col=date)) +
  geom_point() +
  geom_smooth(method = "lm",formula = y~log(x)) +
  geom_point(data=data.frame(twelve_month_pce = last(joined_df$twelve_month_pce),mean_yield = last(joined_df$mean_yield),date = last(joined_df$date)),color="red") +
  scale_x_continuous(breaks = seq(0,10,by = .5)) + 
  scale_y_continuous(limits=c(0,16),breaks = seq(0,16.25,by = 1.25)) +
  labs(x="TTM PCE Inflation Rate",y="Monthly Average 10-Year Yield",title="10-Year Yield by PCE Inflation Rate")
```

## Controlling for time effects

The problem with this plot is that it fails to control for the decrease of 10-year yields over time. Let's investigate whether the time effect on yields is separate from the inflation rate effect. (Spoiler: it is.)

```{r chart_2, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
#Plot yield vs. date
joined_df %>% 
  mutate(twelve_month_pce_strat = factor(round(twelve_month_pce))) %>%
  ggplot(aes(x=date,y=mean_yield)) +
  geom_point() +
  geom_smooth(method="lm",formula=y~log(x)) +
  labs(x="Date",y="Monthly Average 10-Year Yield",title="10-Year Yield over Time")
```

One way to check whether the time effect is separate from the inflation rate effect is to stratify by inflation rate, and then plot yield vs. time for each stratum. Removing strata with fewer than 20 data points, we get four plots that all show a decline in yields over time.

```{r chart_3, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
#Stratify by PCE inflation rate and plot yield vs. date
joined_df %>% 
  mutate(twelve_month_pce_strat = factor(round(twelve_month_pce))) %>%
  group_by(twelve_month_pce_strat) %>%
  mutate(n=n()) %>%
  filter(n>20) %>%
  ggplot(aes(x=date,y=mean_yield)) +
  geom_point() +
  geom_smooth(method="lm",formula=y~log(x)) +
  facet_wrap( ~ twelve_month_pce_strat) +
  labs(x="Date",y="Monthly Average 10-Year Yield",title="10-Year Yield over Time, Stratified by PCE Inflation Rate")
```

Another way to check for an independent time effect is to subtract our PCE inflation-rate regression model's "predicted" yields from every y-value in our data set. Then we plot the leftover "residual" y-values against the date. Here, again, we find a decline in yields over time.

```{r chart_4, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
#Subtract PCE inflation model from the data and regress for time effect
joined_df %>% 
  mutate(pce_model = predict(lm(mean_yield ~ log(twelve_month_pce), data = .)),
         mean_yield_residual = mean_yield - pce_model) %>%
  ggplot(aes(x=date,y=mean_yield_residual)) +
  geom_point() +
  geom_smooth(method="lm",formula=y~x) +
  labs(x="Date",y="Residual Monthly Average 10-Year Yield",title="Residual 10-Year Yield over Time after Subtracting PCE Inflation-Rate\nLog Regression Model")

#Do a multiple linear regression accounting for both PCE inflation and date
fit <- joined_df %>% lm(mean_yield ~ sequence + log(twelve_month_pce), data = .)

#Predict the model for rates today
df <- data.frame(twelve_month_pce = c(seq(1,8.5,by=0.5),last(joined_df$twelve_month_pce)),sequence=rep(530,times=17)) %>%
  mutate(predicted_yield = predict(fit,newdata = .)) %>%
  mutate(highlight = twelve_month_pce %in% last(joined_df$twelve_month_pce))
final_predicted_yield <- predict(fit,newdata=data.frame(twelve_month_pce = last(joined_df$twelve_month_pce),sequence = last(joined_df$sequence)))
```

Running a multiple regression on our data will give us a combined model that incorporates both the PCE inflation rate effect and the time effect. But should we use log regression, linear regression, or some combination of the two? To find out, we use a machine learning method called 10-fold cross-validation. This involves partitioning the data ten times, with 90% of the data in a training set, and the other 10% in a testing set. Then we run regression models on the training data and use them to make predictions on the testing data, and we choose the model that, on average, minimizes the residual mean squared error of predictions across all ten trials.

```{r cross-validation, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

#Partition data for five-fold cross-validation
library(caret)
set.seed(1)
test_index <- createFolds(joined_df$mean_yield,k=10)

#Define function to get residual mean squared error of predictions
RMSE <- function(true_ratings, predicted_ratings){
  sqrt(mean((true_ratings - predicted_ratings)^2))
}

#Perform five-fold cross-validation and calculate RMSEs
rmses <- map_dfr(1:5,function(x){
  #Partition data
  train_set <- joined_df[-test_index[[x]],]
  test_set <- joined_df[test_index[[x]],]
  
  #Do a multiple regression, accounting for both PCE inflation and date
  fit <- train_set %>% lm(mean_yield ~ sequence + twelve_month_pce, data = .)
  fit1 <- train_set %>% lm(mean_yield ~ sequence + log(twelve_month_pce), data = .)
  fit2 <- train_set %>% lm(mean_yield ~ log(sequence) + twelve_month_pce, data = .)
  fit3 <- train_set %>% lm(mean_yield ~ log(sequence) + log(twelve_month_pce), data = .)
  
  #Choose the model that minimizes RMSE
  return(data.frame(Model=c("Double linear","Linear date, log PCE","Log date, linear PCE","Double log"),
                    RMSE = c(RMSE(test_set$mean_yield,predict(fit,test_set)),
                             RMSE(test_set$mean_yield,predict(fit1,test_set)),
                             RMSE(test_set$mean_yield,predict(fit2,test_set)),
                             RMSE(test_set$mean_yield,predict(fit3,test_set))),
                    Trial = rep(x,times=4)))
})

#Choose model that minimizes RMSE on average
rmses %>% group_by(Model) %>% summarize(Mean_RMSE = mean(RMSE)) %>% knitr::kable()

#Retrain the winning model on the whole data set
fit <- joined_df %>% lm(mean_yield ~ sequence + log(twelve_month_pce), data = .)

#Predict the model for rates today
df <- data.frame(twelve_month_pce = c(seq(1,8.5,by=0.5),last(joined_df$twelve_month_pce)),sequence=rep(530,times=17)) %>%
  mutate(predicted_yield = predict(fit,newdata = .)) %>%
  mutate(highlight = twelve_month_pce %in% last(joined_df$twelve_month_pce))
predicted_yield <- predict(fit,newdata=data.frame(twelve_month_pce = last(joined_df$twelve_month_pce),sequence = last(joined_df$sequence)))
```

The winning model uses a linear regression on the date, and a log regression on the mean trimmed PCE inflation rate. We re-train this model on the whole data set.

With this model, we can predict 10-year yields by PCE inflation rate for the current date. At the current level of trimmed-mean PCE inflation, this gives us a predicted 10-year yield of `r paste(round(final_predicted_yield,2),"%",sep="")`. 

```{r chart_5, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
#Plot the rate prediction curve for the current date
df %>% 
  ggplot(aes(x=twelve_month_pce,y=predicted_yield,col=highlight)) +
  geom_point(size=2,show.legend = FALSE) +
  geom_line(show.legend = FALSE) +
  geom_text(data=df[df$highlight==TRUE,],aes(x=twelve_month_pce,y=predicted_yield,label=paste(round(predicted_yield,1),"%",sep=""),nudge_x=-0.4)) +
  scale_x_continuous(breaks=seq(0,10,by=.5)) + 
  labs(x="TTM PCE Inflation Rate",y="Monthly Average 10-Year Yield",title="Predicted 10-Year Yield by PCE Inflation Rate for the Current Date") + 
  theme(legend.position = "none")
```

## Conclusion

Subtracting all time effects prior to the current date, we can re-plot our original plot of 10-year bond yields against historical PCE inflation rates, this time controlling for changes in yields over time. As the chart shows, our model's predicted 10-year yield at about `r paste(round(final_predicted_yield,1),"%",sep="")` is a little below today's roughly `r paste(round(last(tenyr_df$mean_yield),1),"%",sep="")` yield, and far below the approximately `r paste(round(initial_predicted_yield,1),"%",sep="")` yield predicted by a log regression on the PCE inflation rate without controlling for the change in yields over time.

```{r chart_6, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
#Subtract all time effects prior to the present
#Re-plot PCE rate against 10-year yield
joined_df %>%
  mutate(mean_yield = mean_yield - fit$coefficients["sequence"]*sequence + fit$coefficients["sequence"]*last(sequence)) %>%
  ggplot(aes(x=twelve_month_pce,y=mean_yield,col=sequence)) +
  geom_point() +
  geom_smooth(method = "lm",formula = y~log(x)) +
  geom_point(data=data.frame(twelve_month_pce = last(joined_df$twelve_month_pce),mean_yield = last(joined_df$mean_yield),date = last(joined_df$date)),color="red") +
  scale_x_continuous(breaks = seq(0,10,by = .5)) + 
  scale_y_continuous(limits=c(0,16),breaks = seq(0,16.25,by = 1.25)) +
  labs(x="TTM PCE Inflation Rate",y="Monthly Average 10-Year Yield",title="10-Year Yield by PCE Inflation Rate, Controlling for Changes in Yields\nover Time")
```

To be sure, the 10-year yield could go much higher. The dispersion of data on our chart is wide. However, this analysis suggests that the 10-year bond may be fairly priced or even slightly underpriced for the current level of inflation, making this a reasonable time to invest in 10-year bonds.

```{r cleanup, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
#Close Selenium driver
driver$close()
rD[["server"]]$stop()
rm(driver, rD)
gc()
system("taskkill /im java.exe /f", intern=FALSE, ignore.stdout=FALSE)
```

## Bibliography

Board of Governors of the Federal Reserve System (US). 2022a. “Market Yield on u.s. Treasury Securities at 10-Year Constant Maturity, Quoted on an Investment Basis [Dgs10].” FRED. Federal Reserve Bank of St. Louis. https://fred.stlouisfed.org/series/DGS10.

———. 2022b. “Trimmed Mean PCE Inflation Rate [Pcetrim12m159sfrbdal].” FRED. Federal Reserve Bank of St. Louis. https://fred.stlouisfed.org/series/PCETRIM12M159SFRBDAL.
